---
title: 'P1: Describing a Perceptual Phenomenon'
author: "Tobias Weise"
date: "March 30, 2016"
output: 
  html_document: 
    number_sections: no
    theme: journal
    toc: yes
---

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)
setwd("C:/Users/Tobias Weise/Dropbox/data_analyst_nanodegree/P1 - Perceptual Phenomenon")
```

In the following document, I discuss an analysis of a perceptual phenomenon -- the [*Stroop
effect*](https://en.wikipedia.org/wiki/Stroop_effect). It shows that people take longer and 
are more likely to make errors when identifying the ink color of words when those words are 
adjectives describing different colors.  


# Describing the experiment

For the experiment, participants are presented with a list of words displayed in different
ink colors. They are asked to say the color of the ink the word is written in. The individuals
are presented with two kinds of words-ink combinations. First, *congruent words* are those
where the color of ink matches the displayed color adjective. Second, *incongruent words* are
those where the color of ink does not match the displayed color word (e.g. `blue` printed in red).
Each participant gets the same number of congruent and incongruent words. Time for saying
the ink color is recorded for both congruent and incongruent words.


The *dependent variable* of this experiment is the recorded time to say the color of the
ink. *The independent variables* are the kinds of words (congruent or incongruent) that
the test subjects are evaluating.


# Developing a hypothesis and a hypothesis test

Let's assume that the time it takes to say the color of ink is dependent on the kind of words
that test subjects evaluate. Because we know about the Stroop effect, we assume that it will take
individuals longer to evaluate incongruent words than congruent words. With $\mu_{c}$ representing
the mean time for evaluating congruent words and $\mu_{i}$ the mean time for representing
incongruent words, I test the following alternative hypothesis:

$$H_{A}: \mu_{i} > \mu_{c}$$

The null hypothesis is that $\mu_{i}$ and $\mu_{c}$ are equal, thus:

$$H_{0}: \mu_{i} + \mu_{c} = 0$$

To *test the hyptheses*, I run a *dependent t-test for paired samples.* The test is appropriate
because it fits the repeated treatment experimental design: individuals are treated with both
congruent and incongruent words. I use a *one-tailed* version of the test because I assume that
$\mu_{i} > \mu_{c}$. My expectations about the effects of the treatment are directed.


# Descriptive Statistics

First, let's have a look at the general data structure. The following table shows the data.

```{r}
# import data
data <- read.csv("stroopdata.csv")
```

```{r, results = "asis"}
library(googleVis)
op <- options(gvis.plot.tag='chart')
Table <- gvisTable(data, options=list(title="Stroop Effect Histogram"))
plot(Table)
```

Each row represents the values of the recorded time for congruent and incongruent words of 
one individual that took the test. 

Next, here are a number of descriptive statistics of the test results:

-  $n = 24$, i.e. 24 people took the test.
-  $\bar{x_{c}} = `r round(mean(data$Congruent), 2)`$ and $\bar{x_{i}} = `r round(mean(data$Incongruent), 2)`$, 
    i.e. on average, it took the individuals 14.05 seconds to read the congruent words and 22.02 seconds
    to read the incongruent words.
-  $median_{c} = `r round(median(data$Congruent), 2)`$ and $median_{i} = `r round(median(data$Incongruent), 2)`$, 
    i.e. the individuals in the middle of the distribution took 14.36 seconds for reading congruent words
    and 21.02 seconds for reading incongruent words.
-  $\sigma_{c} = `r round(sd(data$Congruent), 2)`$ and $\sigma_{i} = `r round(sd(data$Incongruent), 2)`$, i.e. the standard deviation, and thus the spread of the data, is larger for the incongruent words. 


The summary statistics show that there is considerable difference in the data. This can also be
seen in the following visualizations of the data. First, the following histogram shows the distribution
of the the dependent variable for each group of words. The chart shows that the distribution for the 
incongruent words is further to the right on the x-axis, illustrating the larger mean and median for 
incongruent words and thus that individuals took more time to read the incongruent words. Also, the
larger spread of the incongruent words response time becomes visible.


```{r, results = "asis"}

Histogr <- gvisHistogram(data = data, options=list(title="Histogram: Response Time by Kind of Word", width = "800px", height = "800px"))
plot(Histogr)

```

The same becomes visible in the boxplots chart below. Again, the boxplots show the larger median for
the incongruent words. Further, the boxplots well illustrate that the spread of the response times
is larger for the incongruent words (larger box size, presence of outliers).

```{r}
library(ggplot2)
library(tidyr)
dat <- gather(data, key = experiment, results)
ggplot(dat) + geom_boxplot(aes(y = results, x = experiment)) + theme_bw() + ggtitle("Boxplots: Response Time by Kind of Word")
```

# Hypothesis test and interpretation of results

Next, I present the results of the one-tailed dependent t-test for paired samples:

```{r}
results <- t.test(data$Congruent, data$Incongruent, paired = T, alternative = "less")
results
```

Here are the results in more detail:

-  the point estimate of difference is `r mean(data$Congruent) - mean(data$Incongruent)`. Thus on average,
  people need about 8 seconds more to read incongruent words.
-  the standard deviation of the difference is `r sd(data$Congruent - data$Incongruent)`
-  the standard error of the difference is `r (sd(data$Congruent - data$Incongruent) / sqrt(length(data$Incongruent)))`
-  the t-statistic is `r (mean(data$Congruent) - mean(data$Incongruent)) / ((sd(data$Congruent - data$Incongruent)) / sqrt(length(data$Incongruent)))`
-  the critical values for an alpha level of 0.05 for a one-tailed test is -1.7139
-  the calculated t-value is well below -1.7139. Thus we reject the null-hypothesis $H_{0}$
-  the difference between the groups is statistically significant at a p-value of < 0.0001

Again, I reject $H_{0}$. Thus, the data show that the Stroop effect is indeed at work in the
sample of individuals. Not only does the descriptive statistic show that reading time for 
incongruent pairs is longer. The t-test results also show that this difference is significantly
different with a chance of being caused by change of < 0.001. 


# Digging deeper

Is the data normally distributed? This is a core assumption of the t-test. So, one should test if the
underlying data is normally distributed. Here, I present the results of a Shapiro-Wilk-Test for 
normality.

First, for the group of congruent words, we find that the data appears to be normally distributed:
```{r}
# congruent data
shapiro.test(data$Congruent)
```

However, for the group of incongruent words, we find that they are not:
```{r}
shapiro.test(data$Incongruent)
library(car)
qqp(data$Incongruent, distribution = "norm")
```

As the *q-q-plot* suggests, there are two outliers in the data that might influence results because
they are not covered by the normal distribution. Those are cases `15` and `20` that have extremely
high response times.

Thus, a first strategy could be to exclude those cases and re-run the t-test. Here are the results,
which still confirm the findings of the first t-test. Thus, the two outlier cases do not strongly 
bias the data and the results of the original t-test.

```{r}
# remove outliers
new_data <- data
new_data[15, ] <- NA
new_data[20, ] <- NA
new_data <- na.omit(new_data)

# run the test
t.test(new_data$Congruent, new_data$Incongruent, paired = T, alternative = "less")
```

